Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-16 ind_files/05_unification_discussion.tex
Add some minor discussion on QFT methods

If the `intuitive' version of It\^{o}'s formula for $\mathcal{M}_K(\mathcal{T})$ valued stochastic processes obtained by `taking the limit' $m \to \infty$ in the It\^{o}'s formula for $m$-dimensional stochastic processes holds, then the exact same steps carried out in Appendices \ref{App_density_to_freq} and \ref{App_stoch_var_eqns} will `go through' essentially unchanged for quantitative traits and yield the equations obtained by simply taking $m \to \infty$ in \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} (with sums replaced by integrals and derivatives replaced by functional derivatives as appropriate) as the `fundamental theorems' for the evolution of quantitative traits. Indeed,~\cite{week_white_2021} have recently proposed exactly the It\^o formula we would need via a heuristic It\^o multiplication table for $L^2(\mathbb{R}, m)$ valued processes with a.s. finite Hilbert-Schmidt norm and have shown that their heuristics are equivalent to the rigorous infinite-dimensional stochastic calculus proposed by~\cite{da_prato_stochastic_2014} for more general Hilbert space valued processes\footnote{If this sentence sounds like abstract gibberish to you, don't worry too much about the details - The essence is that~\cite{week_white_2021} have proven the `correct' formula we need holds whenever $b(x|\nu), d(x|\nu)$ and $\mathcal{M}_K(\mathcal{T})$ together fulfill certain technical requirements. My heuristics here are only a `first step' and have focused on accessibility over mathematical propriety, and thus make no attempt to verify whether and when these technical requirements are satisfied.}. Furthermore, using these `spacetime white noise heuristics' together with a particular functional form for (weak) solutions to certain SPDEs,~\cite{week_white_2021} have arrived at precisely the infinite-dimensional version of equations \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} obtained by `taking the limit' $m \to \infty$ in \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} from the `SPDE side' under certain additional assumptions required to ensure existence/uniqueness of solutions to the relevant SPDEs. Equation \eqref{disc_functional_field_eqns} can thus be argued to be the (functional) Fokker-Planck view of the stochastic processes for the evolutionary ecology of quantitative traits recently studied in~\cite{week_white_2021}. Like~\cite{week_white_2021}, I believe that rigorously establishing the precise criteria needed for general existence and uniqueness of solutions to equations of the form \eqref{disc_functional_field_eqns} (or the SPDE versions presented in~\cite{week_white_2021}, which resemble my equation \eqref{functional_langevin}) presents an important mathematical direction for future work that may yield new insights into both the biology of quantitative traits and the mathematics of measure-valued stochastic processes.

More generally, the well-known intimate relation between It\^o SDEs and Fokker-Planck equations (see section \ref{sec_math_background}) is extremely useful, and indeed has been extensively exploited~\citep{van_kampen_stochastic_1981,oksendal_stochastic_1998, gardiner_stochastic_2009}, because many problems are much more easily attacked using probabilistic tools that operate on It\^o process, whereas others are more amenable to study using tools from PDE theory that operate on the Fokker-Planck equations. In the infinite-dimensional case, working with the `physics' side through functional Fokker-Planck equations such as \eqref{disc_functional_field_eqns} allows us to use tools with an essentially `differential equation' flavor such as spectral methods (\cite{rogers_demographic_2012, rogers_modes_2015}; See also Appendix \ref{App_examples} for a general model-independent pedagogical example for studying phenotypic clustering and evolutionary branching), slow manifold (`adiabatic') approximations~\citep{parsons_dimension_2017}, and WKB style expansions~(\cite{rogers_demographic_2012,assaf_wkb_2017}; though it bears noting that WKB theory is closely related to large deviation principles that occur in the study of Markov processes and come in a distinctly `probabilistic' flavor, see for example~\cite{dembo_large_1998}). On the other hand, many biologically interesting questions are much easier to approach from the `mathematics'\footnote{The `mathematics' vs `physics' distinction is just a caricature and is not intended to be read into or thought about too deeply :)} side by working with branching processes/SPDEs and using tools that have an essentially `probabilistic' flavor such as duality~\citep{dawson_stochastic_1975,greenman_duality_2020}, infinitesimal generators~\citep{ethier_markov_1986,etheridge_mathematical_2011}, martingale theory~\citep{champagnat_unifying_2006,etheridge_mathematical_2011}, and weak/mild solutions to SPDEs~\citep{week_white_2021}. Rigorously establishing and exploiting relations between functional Fokker-Planck equations of the form \eqref{disc_functional_field_eqns} and SPDEs of the form \eqref{functional_langevin} (or those studied in~\cite{week_white_2021}) for measure-valued branching processes could thus prove very fruitful for developing more integrative eco-evolutionary theory since it allows us to seamlessly transition between complementary views of the same object to suit the problem at hand.

To the best of my knowledge, a general formulation of stochastic field equations for the population dynamics of quantitative traits from the functional Fokker-Planck side in the manner I have carried out here has not been done before. Similar equations have been formulated for some specific stochastic models of quantitative trait evolution~\citep{rogers_demographic_2012,rogers_modes_2015} and population ecology of size-structured communities~\citep{odwyer_integrative_2009}. Stochastic field equations are also known in mathematical neurobiology~\citep{buice_field-theoretic_2007,bressloff_stochastic_2010,coombes_neural_2014}, and have recently been proposed in a model of collective motion~\citep{o_laighleis_minimal_2018}. Broadly similar deterministic field theoretic approaches have also been proposed to study bio-geography and spatial biodiversity patterns such as the species-area relationship~(\cite{odwyer_field_2010}; But see~\cite{grilli_absence_2012} for an explanation of why the calculations presented in~\cite{odwyer_field_2010} are not exactly correct\footnote{The essential problem is that an assumption about time reversibility (`detailed balance' in more accurate physics lingo), usually made for equilibrium systems, is not satisfied. Nevertheless, the general \emph{approach} is still interesting and seems reasonable}). The ubiquity of such approaches in diverse areas of theoretical biology suggests that the formal systematic study and analysis of such equations could have wide-spread applications.

Currently, (stochastic) field equations are primarily used by physicists working in areas such as statistical field theory and quantum field theory, and are largely attacked using ingenious and often somewhat system-specific tools that may not necessarily be mathematically well understood~(\cite{carmona_stochastic_1999}; but see \cite{bogachev_fokker-planck-kolmogorov_2015} for a rigorous treatment of some infinite-dimensional Fokker-Planck equations). Equation \eqref{disc_functional_field_eqns} opens up the study of quantitative trait dynamics in finite fluctuating populations to analysis using some of these tools - For example, a careful reading of the literature provides many hints and recent attempts illustrating how the major ideas behind the Fock space representation~\citep{del_razo_probabilistic_2022}, perturbative `loop diagrams/expansions'~\citep{dodd_many-body_2009}, and the path integral formalism~\citep{doi_second_1976, peliti_path_1985, chow_path_2015, weber_master_2017} can all be co-opted to coax biological insights from the sort of stochastic processes modeled by \eqref{disc_functional_field_eqns}.

>>>>

If the `intuitive' version of It\^{o}'s formula for $\mathcal{M}_K(\mathcal{T})$ valued stochastic processes obtained by `taking the limit' $m \to \infty$ in the It\^{o}'s formula for $m$-dimensional stochastic processes holds, then the exact same steps carried out in Appendices \ref{App_density_to_freq} and \ref{App_stoch_var_eqns} will `go through' essentially unchanged for quantitative traits and yield the equations obtained by simply taking $m \to \infty$ in \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} (with sums replaced by integrals and derivatives replaced by functional derivatives as appropriate) as the `fundamental theorems' for the evolution of quantitative traits. Indeed,~\cite{week_white_2021} have recently proposed exactly the It\^o formula we would need via a heuristic It\^o multiplication table for $L^2(\mathbb{R}, m)$ valued processes with a.s. finite Hilbert-Schmidt norm and have shown that their heuristics are equivalent to the rigorous infinite-dimensional stochastic calculus proposed by~\cite{da_prato_stochastic_2014} for more general Hilbert space valued processes\footnote{If this sentence sounds like abstract gibberish to you, don't worry too much about the details - The essence is that~\cite{week_white_2021} have proven the `correct' formula we need holds whenever $b(x|\nu), d(x|\nu)$ and $\mathcal{M}_K(\mathcal{T})$ together fulfill certain technical requirements. My heuristics here are only a `first step' and have focused on accessibility over mathematical propriety, and thus make no attempt to verify whether and when these technical requirements are satisfied.}. Furthermore, using these `spacetime white noise heuristics' together with a particular functional form for (weak) solutions to certain SPDEs,~\cite{week_white_2021} have arrived at precisely the infinite-dimensional version of equations \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} obtained by `taking the limit' $m \to \infty$ in \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} from the `SPDE side' under certain additional assumptions required to ensure existence/uniqueness of solutions to the relevant SPDEs. Equation \eqref{disc_functional_field_eqns} is the (functional) Fokker-Planck view of the stochastic processes for the evolutionary ecology of quantitative traits recently studied in~\cite{week_white_2021}. Like~\cite{week_white_2021}, I believe that rigorously establishing the precise criteria needed for general existence and uniqueness of solutions to equations of the form \eqref{disc_functional_field_eqns} (or the SPDE versions presented in~\cite{week_white_2021}) presents an important mathematical direction for future work that may yield new insights into both the biology of quantitative traits and the mathematics of measure-valued stochastic processes.

More generally, the well-known intimate relation between It\^o SDEs and Fokker-Planck equations (see section \ref{sec_math_background}) is extremely useful, and indeed has been extensively exploited~\citep{van_kampen_stochastic_1981,oksendal_stochastic_1998, gardiner_stochastic_2009}, because many problems are much more easily attacked using probabilistic tools that operate on It\^o process, whereas others are more amenable to study using tools from PDE theory that operate on the Fokker-Planck equations. In the infinite-dimensional case, working with the field theory through functional Fokker-Planck equations such as \eqref{disc_functional_field_eqns} allows us to use tools with an essentially `differential equation' flavor such as spectral methods (\cite{rogers_demographic_2012, rogers_modes_2015}; See also Appendix \ref{App_examples} for a general model-independent pedagogical example for studying phenotypic clustering and evolutionary branching), slow manifold (`adiabatic') approximations~\citep{parsons_dimension_2017}, and WKB style expansions~(\cite{rogers_demographic_2012,assaf_wkb_2017}; though it bears noting that WKB theory is closely related to large deviation principles that occur in the study of Markov processes and come in a distinctly `probabilistic' flavor, see~\cite{dembo_large_1998}). On the other hand, many biologically interesting questions are much easier to approach from the branching processes/SPDE side using tools that have an essentially `probabilistic' flavor such as duality~\citep{dawson_stochastic_1975,greenman_duality_2020}, infinitesimal generators~\citep{ethier_markov_1986,etheridge_mathematical_2011}, martingale theory~\citep{champagnat_unifying_2006,etheridge_mathematical_2011}, and weak/mild solutions to SPDEs~\citep{week_white_2021}. Rigorously establishing and exploiting relations between functional Fokker-Planck equations of the form \eqref{disc_functional_field_eqns} and SPDEs of the form \eqref{functional_langevin} (or those studied in~\cite{week_white_2021}) for measure-valued branching processes could thus prove very fruitful for developing more integrative eco-evolutionary theory since it allows us to seamlessly transition between complementary views of the same object to suit the problem at hand\footnote{Indeed, the physics literature has long recognized this. Physicists regularly exploit the connections between stochastic/quantum field equations and SPDEs, often studying SPDEs in \emph{classical} systems using tools and language from quantum field theory~\citep{hochberg_effective_1999,baez_quantum_2018}.}.

To the best of my knowledge, a general formulation of stochastic field equations for the population dynamics of quantitative traits from the functional Fokker-Planck side in the manner I have carried out here has not been done before. Similar equations have been formulated for some specific stochastic models of quantitative trait evolution~\citep{rogers_demographic_2012,rogers_modes_2015} and population ecology of size-structured communities~\citep{odwyer_integrative_2009}. Stochastic field equations are also known in mathematical neurobiology~\citep{buice_field-theoretic_2007,bressloff_stochastic_2010,coombes_neural_2014}, and have recently been proposed in a model of collective motion~\citep{o_laighleis_minimal_2018}. Broadly similar deterministic field theoretic approaches have also been proposed to study bio-geography and spatial biodiversity patterns such as the species-area relationship~(\cite{odwyer_field_2010}; But also see~\cite{grilli_absence_2012} - the actual details of the calculations presented in~\cite{odwyer_field_2010} are not correct). The ubiquity of such approaches in diverse areas of theoretical biology suggests that the formal systematic study and analysis of such equations could have wide-spread applications.

Currently, (stochastic) field equations are primarily used by physicists working in areas such as statistical field theory and quantum field theory, and are largely attacked using ingenious and often somewhat system-specific tools that may not necessarily be mathematically well understood~(\cite{carmona_stochastic_1999}; but see \cite{bogachev_fokker-planck-kolmogorov_2015} for a rigorous treatment of some infinite-dimensional Fokker-Planck equations). Equation \eqref{disc_functional_field_eqns} opens up the study of quantitative trait dynamics in finite fluctuating populations to analysis using some of these tools - For example, a careful reading of the literature provides many hints (albeit couched in physics language) for how (quantum) field theoretic ideas such as the Fock space representation~\citep{del_razo_probabilistic_2022}, `loop diagrams/expansions'~\citep{hochberg_effective_1999,dodd_many-body_2009}, and the often closely related path integral formalism~\citep{doi_second_1976, peliti_path_1985,hochberg_effective_1999, chow_path_2015, weber_master_2017}, can all be co-opted to coax biological insights from the sort of stochastic processes modeled by \eqref{disc_functional_field_eqns}.
